Noncoherent least squares estimators of carrier phase and amplitude

### 2012-10-26 other/confnoncoherent/paper.tex
added some of barry's edits

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-ary phase shift keying.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator is not consistent, it converges to a positive real number that is a function of the true carrier amplitude, the noise distribution, and the size of the constellation.  The results of Monte Carlo simulations are provided and these corroborate the theoretical results.
\end{abstract}
\begin{keywords}
Noncoherent detection, phase shift keying, asymptotic statistics
\end{keywords}

@@

Here we are interested in \emph{noncoherent detection}, where $s_1, \dots, s_L$ are not known at the receiver, and must also be estimated.  This estimation problem has undergone extensive prior study and is often called \emph{multiple symbol differential detection}~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990}.  A practical approach is the least squares estimator,
\begin{equation}\label{eq:hatadef}
\hat{a} = \arg\min_{a \in \complex} \min_{s_1, \dots, s_L \in \calC} \sum_{i = 1}^L \abs{ y_i - a s_i }^2,
\end{equation}
where $\calC$ is the set of symbols from the $M$-PSK constellation.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is white and Gaussian.  As we will show, the estimator can work well even when the noise is not Gaussian.  Mackenthun~\cite{Mackenthun1994} described an algorithm to compute the least squares estimator $\hat{a}$ that requires only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.

@@

In the literature it has been common to assume that the symbols $s_1, \dots, s_L$ are of primary interest and the complex amplitude $a_0$ is a nuisance parameter.  The metric of performance is correspondingly \emph{symbol error rate}, or \emph{bit error rate}.  Whilst estimating the symbols (or more precisely the transmitted bits) is ultimately the goal, we take the opposite approach here.  Our aim is to estimate $a_0$, and we treat the unknown symbols as nuisance parameters.  This is motivated by the fact that in many modern communication systems the data symbols are \emph{coded}.  For this reason raw symbol error rate is not of interest at this stage.  Instead, we desire an accurate estimator $\hat{a}$ of $a_0$, so that the compensated received symbols $\hat{a}^{-1}y_i$ can be accurately modelled using an additive noise channel.  The additive noise channel is a common assumption for subsequent receiver operations, such as decoding.  The estimator $\hat{a}$ is also used in the computation of decoder metrics for modern decoders, and for interference cancellation in multiuser systems.  Consequently, our metric of performance will not be symbol or bit error rate, it will be $\abs{a_0 - \hat{a}}^2$. It will actually be informative to consider the carrier phase and amplitude estimators separately, that is, if $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{\rho}$ is a positive real number, then we consider $\langle\theta_0 - \hat{\theta}\rangle^2$ and $(\rho_0 - \hat{\rho})^2$.  The function $\fracpart{\cdot}$ denotes its argument taken `modulo $\tfrac{2\pi}{M}$' into the interval $[-\pi/M, \pi/M)$.  It will become apparent why $\langle\theta_0 - \hat{\theta}\rangle^2$ rather than $(\theta_0 - \hat{\theta})^2$ is the appropriate measure of error for the phase parameter.

@@

The next two theorems describe the asymptotic properties of the least squares estimator.  We omit the proofs due to space constraints.  Proofs will be provided in an upcoming paper.

>>>>

\begin{abstract}
We consider least squares estimators of carrier phase and amplitude from a noisy communications signal.  We focus on signaling constellations that have symbols evenly distributed on the complex unit circle, i.e., $M$-ary phase shift keying.  We show, under reasonably mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent and asymptotically normally distributed.  However, the amplitude estimator is not consistent, but converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation.  The results of Monte Carlo simulations are provided and these corroborate the theoretical results.
\end{abstract}
\begin{keywords}
Noncoherent detection, phase shift keying, asymptotic statistics
\end{keywords}

@@

In the paper we are interested in \emph{noncoherent detection}, where $s_1, \dots, s_L$ are not known at the receiver, and must also be estimated.  This estimation problem has undergone extensive prior study and is often called \emph{multiple symbol differential detection}~\cite{ViterbiViterbi_phase_est_1983,Cowley_ref_sym_carr_1998,Wilson1989,Makrakis1990,Liu1991,Mackenthun1994,Sweldens2001,McKilliamLinearTimeBlockPSK2009,Divsalar1990}.  A practical approach is the least squares estimator,
\begin{equation}\label{eq:hatadef}
\hat{a} = \arg\min_{a \in \complex} \min_{s_1, \dots, s_L \in \calC} \sum_{i = 1}^L \abs{ y_i - a s_i }^2,
\end{equation}
where $\calC$ is the set of symbols from the $M$-PSK constellation.  The least squares estimator is also the maximum likelihood estimator under the assumption that the noise sequence $\{w_i, i \in \ints\}$ is white and Gaussian.  As we show, the estimator can work well even when the noise is not Gaussian.  Mackenthun~\cite{Mackenthun1994} described an algorithm to compute the least squares estimator $\hat{a}$ that requires only $O(L \log L)$ arithmetic operations.  Sweldens~\cite{Sweldens2001} rediscovered Mackenthun's algorithm in 2001.

@@

In the literature it has been common to assume that the symbols $s_1, \dots, s_L$ are of primary interest and the complex amplitude $a_0$ is a nuisance parameter.  The metric of performance is correspondingly the \emph{symbol error rate}, or \emph{bit error rate}.  While estimating the symbols (or more precisely the transmitted bits) is ultimately the goal, we take the opposite point of view here.  Our aim is to estimate $a_0$, and we treat the unknown symbols as nuisance parameters.  This is motivated by the fact that in many modern communication systems the data symbols are \emph{coded}.  For this reason raw symbol error rate is not of interest at this stage.  Instead, we desire an accurate estimator $\hat{a}$ of $a_0$, so that the compensated received symbols $\hat{a}^{-1}y_i$ can be accurately modelled using an additive noise channel.  The additive noise channel is a common assumption for subsequent receiver operations, such as decoding.  The estimator $\hat{a}$ is also used in the computation of decoder metrics for modern decoders, and for interference cancellation in multiuser systems.  Consequently, our metric of performance will not be symbol or bit error rate, but $\abs{a_0 - \hat{a}}^2$. It will be informative to consider the carrier phase and amplitude estimators separately, that is, if $\hat{a} = \hat{\rho}e^{j\hat{\theta}}$ where $\hat{\rho}$ is a positive real number, then we consider $\langle\theta_0 - \hat{\theta}\rangle^2$ and $(\rho_0 - \hat{\rho})^2$.  The function $\fracpart{\cdot}$ denotes its argument taken `modulo $\tfrac{2\pi}{M}$' into the interval $[-\pi/M, \pi/M)$, that is
\[
\fracpart{x} = x - \tfrac{2\pi}{M}\operatorname{round}\left(\tfrac{M}{2\pi}x\right),
\]  
where $\operatorname{round}(\cdot)$ takes its argument to the nearest integer.  The direction of rounding for half-integers is not important so long as it is consistent.  We have chosen to round up half-integers here.  It will become apparent why $\langle\theta_0 - \hat{\theta}\rangle^2$ rather than $(\theta_0 - \hat{\theta})^2$ is the appropriate measure of error for the phase parameter.

@@

The next two theorems describe the asymptotic properties of the least squares estimator.  We omit the proofs due to space constraints.  Proofs will be provided in an forthcomming paper.
